Analysing Melbourne CBD pedestrian counts data with a focus on imputation methods

### 2017-10-22 thesis_draft/00-abstract.Rmd
spell check

The City of Melbourne provides an open data platform to access council datasets, with the intention "to increase transparency, improve public services and support new economic and social initiatives". This paper will focus primarily on the pedestrian data collected from pedestrian sensors placed throughout Melbourne's CBD. In particular, we want to be able to model the pedestrian traffic at different locations in the CBD, and forecast the traffic. In addition, we aim to have an interactive, dynamic data visualisation of the model to help us to understand how Melbourne operates.

>>>>

The City of Melbourne provides access pedestrian data collected from pedestrian sensors placed throughout Melbourne's CBD. Through data analysis and exploration, the aim is analyse the pedestrian behaviour at different locations in the CBD in order to predict pedestrian counts. Issues in the dataset, with missing values and false zero counts, are investigated and solutions proposed. An improved imputation method is proposed, using different models based on the proportion of missing values. For sensor locations with a small proportion of missing values used generalised linear models using only time and date based variables as predictors, while sensors with a large proportion of missings used neighbouring sensor counts as predictors. Using this imputed data, predictive models can be trained more effectively, producing better estimates. A predictive model for on-the-fly predictions was wrapped into a function to facilitate point estimates in an easy and tidy format.

### 2017-10-22 thesis_draft/01-chap1.Rmd
spell check

dfa$DayType <-       ifelse(dfa$Day == "Sunday", "Sunday",
                     ifelse(dfa$Day == "Saturday", "Saturday",
                    ifelse(dfa$Day == "Monday", "Monday",
                     ifelse(dfa$Day == "Friday", "Friday",
                                  "Midweek"))))
dfa$DayType[dfa$IsHDay == 1] <- "Holiday"
dfa$HDay <- ifelse(dfa$IsHDay == 1, "Holiday", dfa$Day)
original_df <- dfa
dfa <- filter(dfa, Date_Time < "2017-01-01")
ori_2 <- filter(original_df, Date_Time < "2017-01-01")
ori_ts <- filter(original_df, Date_Time > "2016-12-31")
ori_ts2 <- ori_ts
```
# Introduction

@@

@@

@@

For the purposes of that paper, it was found to be inappropriate with the data being analysed.  However, unlike the highway traffic data, where patterns were compared to traffic volumes in different jurisdictions (large geographical distance), the geographic distances between the sensor locations in the Melbourne CBD pedestrian data are much smaller. Another major difference is the randomness of the missing values, where the periods of missing data run longer in the Melbourne CBD pedestrian data compared to that in the Alberta traffic data. Roh et al. proposed using non-parametric estimation methods with k-Nearest Neighbours (kNN) for imputation, although this is unlikely to work well with the lack of data at certain sensor locations in the Melbourne CBD Pedestrian data. Similarly, most other applications of imputation methods on traffic data used non-parametric regressions for traffic flow forecasting (Zhong, J., Ling, S. 2015).

Another important part of analysis which needed to be conducted when performing data imputation is assessment of robustness (Zhong, M. et al. 2005). Zhong et al. found that modeles using additional observations as input and more sophisticated prediction techniques were able to make better imputations on a consistent basis.

@@

```{r naprop_hist, fig.width=6, fig.height=3, fig.cap=paste("Histograms of the proportions of missing values at each sensor during 2014 - 2016 when calculated before and after correction for false zero counts. The distributions of proportion of missing values at each sensor is significantly different after correcting for false zero counts in the data. This emphasises the necessity to correct for false zero counts as it will cause biased estimates.")}
theme_g <- theme_minimal() +
           theme(plot.title = element_text(hjust = 0.5,
                                           size = 10))
p1 <- ggplot() + geom_histogram(aes(x = na_prop*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
p2 <- ggplot() + geom_histogram(aes(x = na_prop_pre*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
grid.arrange(p2, p1, nrow = 1)
```

@@

Instead of `HDay`, `DayType` is used which classifies Tuesday, Wednesday and Thursday as a single factor level, Midweek. This can be interpreted as the type of day. This can be done as the daily patterns on the Mideweek weekdays are similar to each other. The intuition for the use of this variable is for the reduction in the number of parameters to estimate ($23 \times 2 = 46$ less parameters). Another consequence of this is more robust estimates for the midweek days, being less sensitive to outliers on a particular day. Because this model will be estimated for sensor locations with few missing values,  enough data is available to add in month as an additive effect tol help capture the annual seasonality. This is not possible at locations with a large proportion of missing values, particularly at newer installations where less than 12 months of data is available in the training period specified.

@@

The amount of people passing through the neighbouring sensors will be expected to be a strong predictor of the pedestrian counts as they are likely to also pass through. Using geographical neighbours, it is possible capture rare events effectively as the predictors are also random variables. Any large shocks to the counts at the neighbouring sensors will have an effect on the predicted counts.

@@

A for loop is used to estimate the GLM specified:
```{r, echo = TRUE, eval = FALSE}
glm(dfa[, i] ~ Month + DayType*Time, data = dfa,
    family = quasipoisson())
```
where `dfa` is the object name given to the data frame containing the training data after the false zero-count correction, and `i` is the $i^{th{}}$ sensor in the list of sensors with a small proportion of missing values. These models are stored in a list, forming a list of models. An alternative approach to using a for loop in R is splitting the data into separate data frames for each sensor location (forming a list of data frames) and applying the `glm()` function on each data frame in the list using the function `purrr::map()`. However, when this was attempted, performance issues were encountered due to the large size of the estimated model objects in R. This was not an issue when a for loop was used. Additional code for removing unnecessary elements of the GLM model object before storing to the list of models, reduucing the memory usage.

@@

A seperate MARE for predictions in-sample (in the training set) and the predictions out of sample (test set of 2017 data) is calculated. This is to help identify overfitting of the training data.

@@

More importantly, the performance of the model's predictions of the out of sample observations (2017 data) is evavaluated as cross validation. Due to potential false zero counts in the test data sourced from `rwalkr::walk_melb()`, correction of false zero-counts are required to be able to properly evaluate the goodness of fit of out-of-sample observations. The method used for this is identical to that used on the training set in the algorithm, with replacement of repeated value sequences of length 6 hours or longer with the value `NA`.

@@

As these sensor locations were both classified as locations with a large proportion of missing values, it would appear that the relationship between these locations and their neighbouring sensors has changed. This is because the in-saple MARE at these sensor locations were quite acceptable at `r round(MARE_tr["Spencer St-Collins St (North)"]*100, 2)`%, `r round(MARE_tr["Spencer St-Collins St (South)"]*100, 2)`% and `r round(MARE_tr["Flinders St-Swanston St (West)"]*100, 2)`% respectively.

@@

The plot of the average counts by time of day and type of day and data set shows that this location saw an increase in counts on weekdays which did not occur at the neghbouring sensors. It was also seen that the pattern is slightly different during weekdays, with a higher afternoon/evening peak. These types of change in pedestrian traffic behaviour is hard to capture in our imputation model, especially when its neighbours did not have a similar change. While it is possible to add a variable for year to allow for year to year growth, the short period of training data doesn't make this appropriate at this stage.

Another method to perform cross-validation of the robustness of the model is to remove data at a sensor, and evaulate the predictions made. This will allow us to evaluate the imputation model's performance of out of sample but within training period observations, which is the primary concern with them imputation process.  
The sensor selected with a small proportion of missing values, Bourke Street Mall (South), has `r round((1 - na_prop["Bourke Street Mall (South)"])*100, 3)`% of data available. As observed in the data, missing data is present in different ways: random, short periods of missing data (no more that 6 hours) or long periods of missing data (multiple months). Simulation of both types of missingness at this sensor location is performed.

>>>>

dfa$DayType <-       ifelse(dfa$Day == "Sunday", "Sunday",
                     ifelse(dfa$Day == "Saturday", "Saturday",
                    ifelse(dfa$Day == "Monday", "Monday",
                     ifelse(dfa$Day == "Friday", "Friday",
                                  "Midweek"))))
dfa$DayType[dfa$IsHDay == 1] <- "Holiday"
dfa$HDay <- ifelse(dfa$IsHDay == 1, "Holiday", dfa$Day)
original_df <- dfa
dfa <- filter(dfa, Date_Time < "2017-01-01")
ori_2 <- filter(original_df, Date_Time < "2017-01-01")
ori_ts <- filter(original_df, Date_Time > "2016-12-31")
ori_ts2 <- ori_ts

ped_loc <- read.csv("../data/Pedestrian_sensor_locations.csv")
```
# Introduction

@@

```{r, fig.cap=paste("Plot of hourly counts for Flagstaff Station, Southbank and State Library sensors during the training period of 2014-2016. It is observed that Southbank has a period without any data (pre installation), while Flagstaff Station appears to have complete data. State Library appears complete, although the first half of 2015 shows suspeciously low hourly counts for a long period. ")}
pd <- ggplot(dfa %>% filter(Date_Time < "2017-01-01 00:00:00",
                            Date_Time > "2013-12-31 00:00:00")) +
        scale_y_continuous(limits = c(0, 7000)) +
        xlab(NULL) +
        theme_minimal()

p1 <-  pd + geom_line(aes(x = Date_Time, y = `Flagstaff Station`))
p2 <-  pd + geom_line(aes(x = Date_Time, y = `Southbank`))
p3 <-  pd + geom_line(aes(x = Date_Time, y = `State Library`)) +
            xlab("Time and Date")

grid.arrange(p2, p1, p3, nrow = 3, ncol = 1)
```

@@

```{r, fig.cap = paste("Map of all the pedestrian count sensors currently installed around the Melbourne CBD. The proximity between sensors is very close at some locations. This indicates potential for imputation using nearby sensor counts as predictors.")}
melb <- get_map(location=c(mean(range(ped_loc$Longitude)),
                               mean(range(ped_loc$Latitude))),
                                zoom=14)
ggmap(melb) + geom_point(aes(x = Longitude, y = Latitude),
                         data = ped_loc, colour = "red",
                         size = 4, alpha = 0.7, shape = 17) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank())
```

@@

For the purposes of that paper, it was found to be inappropriate with the data being analysed.  However, unlike the highway traffic data, where patterns were compared to traffic volumes in different jurisdictions (large geographical distance), the geographic distances between the sensor locations in the Melbourne CBD pedestrian data are much smaller. Another major difference is the randomness of the missing values, where the periods of missing data run longer in the Melbourne CBD pedestrian data compared to that in the Alberta traffic data. Roh et al. proposed using non-parametric estimation methods with k-Nearest Neighbours (kNN) for imputation, although this is unlikely to work well with the lack of data at certain sensor locations in the Melbourne CBD Pedestrian data. This is because kNN regression works best when there is a lot of historical data. The lack of historical data at newer sensor installations, where less than 12 months of training data is available, makes kNN unfeasible. Similarly, most other applications of imputation methods on traffic data used non-parametric regressions for traffic flow forecasting (Zhong, J., Ling, S. 2015) involving kNN.

Another important part of analysis which needed to be conducted when performing data imputation is assessment of robustness (Zhong, M. et al. 2005). Zhong et al. found that models using additional observations as input and more sophisticated prediction techniques were able to make better imputations on a consistent basis.

@@

```{r, fig.width=6, fig.height=3, fig.cap=paste("Histograms of the proportions of missing values at each sensor during 2014 - 2016 when calculated before and after correction for false zero counts. The distributions of proportion of missing values at each sensor is significantly different after correcting for false zero counts in the data. This emphasises the necessity to correct for false zero counts as it will cause biased estimates.")}
theme_g <- theme_minimal() +
           theme(plot.title = element_text(hjust = 0.5,
                                           size = 10))
p1 <- ggplot() + geom_histogram(aes(x = na_prop*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
p2 <- ggplot() + geom_histogram(aes(x = na_prop_pre*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
grid.arrange(p2, p1, nrow = 1)
```

@@

Instead of `HDay`, `DayType` is used which classifies Tuesday, Wednesday and Thursday as a single factor level, Midweek. This can be interpreted as the type of day. This can be done as the daily patterns on the Midweek weekdays are similar to each other. The intuition for the use of this variable is for the reduction in the number of parameters to estimate ($23 \times 2 = 46$ less parameters). Another consequence of this is more robust estimates for the midweek days, being less sensitive to outliers on a particular day. Because this model will be estimated for sensor locations with few missing values,  enough data is available to add in month as an additive effect to help capture the annual seasonality. This is not possible at locations with a large proportion of missing values, particularly at newer installations where less than 12 months of data is available in the training period specified.

@@

The amount of people passing through the neighbouring sensors will be expected to be a strong predictor of the pedestrian counts as they are likely to also pass through. Using geographical neighbours, it is possible capture rare events effectively as the predictors are also random variables. Any large shocks to the counts at the neighbouring sensors will have an effect on the predicted counts. This method is a form of pattern matching, where the sensor counts being imputed can be considered the study curve, and the neighbouring sensor counts are candidates for matching.

@@

A for loop is used to estimate the GLM specified:
```{r, echo = TRUE, eval = FALSE}
glm(dfa[, i] ~ Month + DayType*Time, data = dfa,
    family = quasipoisson())
```
where `dfa` is the object name given to the data frame containing the training data after the false zero-count correction, and `i` is the $i^{th{}}$ sensor in the list of sensors with a small proportion of missing values. These models are stored in a list, forming a list of models. An alternative approach to using a for loop in R is splitting the data into separate data frames for each sensor location (forming a list of data frames) and applying the `glm()` function on each data frame in the list using the function `purrr::map()`. However, when this was attempted, performance issues were encountered due to the large size of the estimated model objects in R. This was not an issue when a for loop was used. Additional code for removing unnecessary elements of the GLM model object before storing to the list of models, reducing the memory usage. Significant improvements in computing time is made with the use of parallel computing. Parallel computing is implemented with the use of `foreach()` and `doSNOW` rather than using the base R `for()` function. IT was found that estimation time was reduced by 50% when using a cluster size of two parallel nodes. Additional nodes (more than two) did not result in a significantly reduced computing time, so only two were used.

@@

A separate MARE for predictions in-sample (in the training set) and the predictions out of sample (test set of 2017 data) is calculated. This is to help identify overfitting of the training data.

@@

More importantly, the performance of the model's predictions of the out of sample observations (2017 data) is evaluated as cross validation. Due to potential false zero counts in the test data sourced from `rwalkr::walk_melb()`, correction of false zero-counts are required to be able to properly evaluate the goodness of fit of out-of-sample observations. The method used for this is identical to that used on the training set in the algorithm, with replacement of repeated value sequences of length 6 hours or longer with the value `NA`.

@@

As these sensor locations were both classified as locations with a large proportion of missing values, it would appear that the relationship between these locations and their neighbouring sensors has changed. This is because the in-sample MARE at these sensor locations were quite acceptable at `r round(MARE_tr["Spencer St-Collins St (North)"]*100, 2)`%, `r round(MARE_tr["Spencer St-Collins St (South)"]*100, 2)`% and `r round(MARE_tr["Flinders St-Swanston St (West)"]*100, 2)`% respectively.

@@

The plot of the average counts by time of day and type of day and data set shows that this location saw an increase in counts on weekdays which did not occur at the neighbouring sensors. It was also seen that the pattern is slightly different during weekdays, with a higher afternoon/evening peak. These types of change in pedestrian traffic behaviour is hard to capture in our imputation model, especially when its neighbours did not have a similar change. While it is possible to add a variable for year to allow for year to year growth, the short period of training data doesn't make this appropriate at this stage.

Another method to perform cross-validation of the robustness of the model is to remove data at a sensor, and evaluate the predictions made. This will allow us to evaluate the imputation model's performance of out of sample but within training period observations, which is the primary concern with them imputation process.  
The sensor selected with a small proportion of missing values, Bourke Street Mall (South), has `r round((1 - na_prop["Bourke Street Mall (South)"])*100, 3)`% of data available. As observed in the data, missing data is present in different ways: random, short periods of missing data (no more that 6 hours) or long periods of missing data (multiple months). Simulation of both types of missingness at this sensor location is performed.

### 2017-10-22 thesis_draft/02-chap2.Rmd
spell check

Using a GLM predicts well compared to estimating with other models such as AR($p$) or other time series models when considering the tradeoff between model complexity and prediction accuracy. Due to multiple seasonalities, AR($p$) models would need to be considered on a hour of the week basis at each sensor location. Prediction with time series models would also require of historical data. For the objective of this predictive model, the additional imputs required by end users to predict pedestrian counts when making out of sample predictions is not desirable.

@@

>>>>

Using a GLM predicts well compared to estimating with other models such as AR($p$) or other time series models when considering the trade off between model complexity and prediction accuracy. Due to multiple seasonalities, AR($p$) models would need to be considered on a hour of the week basis at each sensor location. Prediction with time series models would also require of historical data. For the objective of this predictive model, the additional inputs required by end users to predict pedestrian counts when making out of sample predictions is not desirable.

@@

Further research to improve the imputation algorithm can be made, such as more alternative models for different proportions of missing values. Cross validation of different missing proportion threshold values could also improve the prediction accuracy of the imputation models.
